---
title: "A Brief Introduction to ERGMs"
subtitle: "SSC 2026 ERGM Workshop · Module 0"
format:
  revealjs:
    theme: [simple, theme.scss]
    slide-number: true
    fig-width: 7
    fig-height: 3.6
execute:
  eval: true    # renv library is healthy; chunks run and bake in outputs + the GOF plot
  echo: true
  warning: false
  message: false
  cache: true
---

## Why this intro

- **No ERGM experience needed**
- Core ideas in 30 minutes, on a familiar example: **teen friendships**
- Builds to the workshop's key concept: **network targets**

::: notes
Everything here runs on `faux.magnolia.high`, which ships with `statnet`, so no workshop
data is required. Module 1 then applies the same ideas to syringe-sharing networks.
:::

---

## The question

**Micro-level processes → macro-level structure**

- How do adolescents form friendships?
- Can we name the processes that generate the patterns we see?

::: notes
Sometimes we care about people's attributes; here the *pattern of ties* is the point.
We'll pin down three things: what a generative process is, what "structure" is, and how to
tell if a process reproduces the structure.
:::

---

## Where we're going

1. **Describe** the network: what does the structure look like?
2. **Summarize** it as numbers: these become our **targets**
3. **Model** it: null vs. assortative mixing
4. **Simulate**: does the model reproduce the targets?

::: notes
This arc (describe, summarize, model, check) is the workshop in miniature.
:::

---

## The data

```{r load}
library(statnet)
data("faux.magnolia.high")
fmh <- faux.magnolia.high

c(students = network.size(fmh),
  ties     = network.edgecount(fmh))
```

Add Health friendships; directionality ignored for illustration.

::: notes
1,461 students, 974 undirected friendship ties. A to B and B to A are the same tie here.
:::

---

## What does it look like?

```{r netplot}
#| echo: false
#| fig-width: 9
#| fig-height: 5.8
#| out-width: "100%"
#| fig-align: center
set.seed(42)
grade <- as.factor(fmh %v% "Grade")
pal   <- hcl.colors(nlevels(grade), "Dark 3")
par(mar = c(0, 0, 0, 0))
plot(fmh,
     vertex.col    = pal[grade],
     vertex.border = pal[grade],
     vertex.cex    = 0.7,
     edge.col      = "#b9c6cc55")
legend("topleft", legend = levels(grade), col = pal,
       pch = 19, bty = "n", cex = 0.7, title = "Grade")
```

::: notes
Colored by grade: ties visibly cluster within grade, with a fringe of isolates. Each point
is a student; lines are friendships. The next two slides put numbers on both.
:::

---

## Who's in the network?

```{r attrs}
table(fmh %v% "Grade")
table(fmh %v% "Sex")
```

::: notes
Each node carries attributes: Grade, Sex, Race. These are the ingredients for the
"social processes" we'll hypothesize.
:::

---

## Structure: who connects with whom?

```{r mixing}
mixingmatrix(fmh, "Sex")
```

Most ties sit on the diagonal: friendships cluster **within** sex.

::: notes
A mixing matrix counts ties by the sex of the two people. Heavy diagonal means assortative
mixing. Same story for grade and race.
:::

---

## Structure: how many friends?

```{r degree}
summary(fmh ~ degree(0:5))
```

Most students have **few** ties; a handful have many. A skewed degree
distribution.

::: notes
`degree(0:5)` counts nodes with exactly 0, 1, ... 5 ties. This shape (many low-degree,
few high-degree) is typical of social networks and will matter for model fit.
:::

---

## These summaries *are* network statistics

```{r targets}
summary(fmh ~ edges +
          nodematch("Grade") +
          nodematch("Race") +
          nodematch("Sex"))
```

Each number counts a feature of the observed network.

::: notes
`edges` is total ties; each `nodematch` is ties within that attribute. We just computed
four numbers that describe the structure. Hold onto them: they're about to become
*targets*.
:::

---

## What is a "network target"?

::: {.punch}
A target is a summary statistic we want the model to reproduce.
:::

- `summary(net ~ terms)` **computes** them from data
- `ergm(net ~ terms, target.stats = …)` **fits a model to those numbers**
- So we can model a network with **only aggregate summaries**, no full
  who-knows-whom matrix

::: notes
This is the crux of the workshop. Here we have the whole network, so the targets are the
observed counts. In Module 1, the targets come from meta-analysis summaries (mixing
matrices, degree distributions) and we never see one full observed network. ERGMs let us
fit to the summaries regardless.
:::

---

## What is an ERGM?

**Logistic regression for ties.**

- **Outcome:** does a tie exist between two people? (0/1)
- **Predictors:** network statistics (the target terms)
- **Coefficients:** log-odds of a tie, converted to probabilities

::: notes
The left-hand side is the observed network; the right-hand side is the terms, which are
exactly the statistics we just computed. R and statnet run the estimation for us. Everything
that follows is this one idea.
:::

---

## Null model: ties form at random

```{r null}
random.m <- ergm(fmh ~ edges)
plogis(coef(random.m))   # baseline tie probability
```

`edges` is the intercept: baseline density. Networks are overwhelmingly
**non**-ties.

::: notes
Log-odds about −6.99, probability ≈ 0.0009. A random pair almost never shares a tie.
:::

---

## Assortative mixing

```{r assort}
assort.m <- ergm(fmh ~ edges +
                   nodematch("Grade") +
                   nodematch("Race") +
                   nodematch("Sex"))
round(coef(assort.m), 2)
```

::: notes
All terms here are dyad-independent, so this fits instantly via logistic regression, no
MCMC. Positive coefficients mean ties are more likely within that attribute.
:::

---

## Interpretation

- Baseline log-odds ≈ −10.0 → probability 4.5 × 10⁻⁵
- Same grade: −10.0 + 3.2 ≈ −6.8 → probability ≈ 0.001

::: {.punch}
A same-grade tie is ~22× more likely than a random pair.
:::

::: notes
The `edges` term is the intercept; each `nodematch` shifts the log-odds when two people
share that attribute. Log-likelihood also jumps (−7790 to −6528): a better fit.
:::

---

## Is a better-fitting model *good enough*?

- A higher log-likelihood says "better"
- But does the model *reproduce* the structure we care about?
- The test: simulate many networks, compare to the data

::: notes
A converged model is not necessarily a well-aligned one. We pick a structural property we
care about and check whether the data sits inside the simulated spread.
:::

---

## Goodness of fit: degree

```{r gof}
#| echo: false
#| fig-width: 8.5
#| fig-height: 5.2
#| out-width: "92%"
#| fig-align: center
plot(gof(assort.m ~ degree))
```

::: notes
Observed degree distribution (line) vs. 100 simulated networks (boxplots). Both models so
far miss part of the structure: the observed line drifts outside the simulated spread.
:::

---

## Observed vs. simulated

```{r obs-sim}
#| echo: false
#| fig-width: 11
#| fig-height: 5.4
#| out-width: "100%"
#| fig-align: center
set.seed(7)
sim   <- simulate(assort.m, nsim = 1)
grade <- as.factor(fmh %v% "Grade")
pal   <- hcl.colors(nlevels(grade), "Dark 3")

op <- par(mfrow = c(1, 2), mar = c(0, 0, 2, 0))
plot(fmh, vertex.col = pal[grade], vertex.border = pal[grade],
     vertex.cex = 0.6, edge.col = "#b9c6cc55", main = "Observed")
plot(sim, vertex.col = pal[grade], vertex.border = pal[grade],
     vertex.cex = 0.6, edge.col = "#b9c6cc55", main = "Simulated")
par(op)
```

::: notes
The fitted model regenerates the within-grade clustering: same generative signature. Same
node set and attributes; only the edges are re-drawn. It captures grade homophily and
density, but (as the GOF showed) still misses higher-order structure like triadic closure.
:::

---

## What's missing?

- Friends of friends tend to connect: **triadic closure**
- Captured with `gwesp(...)` → fit improves further
- Raw triangle counts can cause **degeneracy** (revisited in Module 4)

::: notes
GWESP, the geometrically-weighted edgewise shared partner term, down-weights each
additional shared friend, capturing the non-linear way closure works.
:::

---

## Network targets, revisited

::: {.punch}
Targets are the bridge from data to model.
:::

- Every summary in this module (mixing, degree) is a candidate **target**
- `target.stats` fits the model to those numbers
- Simulation checks whether the model reproduces them
- **Module 1:** targets come from aggregate data, no full network needed

::: notes
The one idea to carry out of this module. We computed targets from a full network here; the
workshop's real setting is fitting to summary targets alone.
:::

---

## Takeaway

::: {.punch}
Describe → summarize as targets → model → simulate to check.
:::

Same machinery in **Module 1**: syringe-sharing among people who inject drugs.

::: notes
That shift, fitting to summary targets rather than one observed network, is exactly why
simulation-as-diagnostic (Module 5) replaces standard `gof()`. Further reading:
Goodreau (2008); Robins et al. (2007).
:::
