---
title: "Module 1: From Aggregate Summaries to Network Parameters"
subtitle: "SSC 2026 ERGM Workshop"
format: revealjs
---

## What is social network analysis?

Taking a **relational perspective**

- People influence each other
- People who share attributes are more likely to be connected
- When studying health, what is happening *around people* needs explicit attention

In this workshop: syringe-sharing networks among people who inject drugs (PWID)
in the Chicago metro area

---

## The data

**n = 1,000 nodes** — a synthetic population mirroring the HepCEP PWID network

**Attributes:** sex, race/ethnicity, age, geographic location (zip code, lat/lon)

**Ties:** directed syringe-sharing partnerships

```{r setup}
library(network)
library(ergm)
library(dplyr)
library(readr)
library(ggplot2)
library(here)

nodes   <- read_csv(here("data", "synthetic", "nodes.csv"))
targets <- readRDS(here("data", "synthetic", "targets.rds"))
```

---

## Inspect the population

```{r inspect}
glimpse(nodes)
table(nodes$sex)
table(nodes$race)
mean(nodes$young)        # proportion age < 26
mean(nodes$chicago == 1) # proportion in Chicago zip codes
```

---

## What is an ERGM?

ERGMs model **which ties exist** as a function of node attributes and network structure

- Operationalized like a logistic regression
- **Left-hand side:** the observed (or target) network
- **Right-hand side:** terms capturing social processes
- **Coefficients:** log-odds of a tie existing

$$\text{logit}(P(y_{ij} = 1)) = \theta_1 g_1(y, X) + \theta_2 g_2(y, X) + \ldots$$

The `edges` term is analogous to the intercept.

---

## Social processes → ERGM terms

| Social process | ERGM term | What it captures |
|----------------|-----------|-----------------|
| Overall density | `edges` | Baseline tie probability |
| Sex mixing | `nodemix("sex")` | Who shares with whom by sex |
| Age mixing | `nodemix("young")` | Young vs. older mixing |
| Racial/ethnic mixing | `nodemix("race.num")` | Within- vs. cross-race ties |
| Degree distribution | `odegree`, `idegree` | How many partners each person has |
| Geography | `dnf(...)` | Proximity shapes tie formation |

---

## Our research question

**Can we identify social processes that generate the observed syringe-sharing
structure in this population?**

This is the question we'll answer — step by step — across this workshop.

---

## Exploring the network structure

```{r net-init}
# Initialize a directed network with node attributes
n0 <- network.initialize(n = nrow(nodes), directed = TRUE)
set.vertex.attribute(n0, colnames(nodes), nodes)
n0
```

---

## Degree distribution: what are we trying to match?

```{r degree-targets}
# Out-degree targets from empirical data (scaled to n=1000)
targets$outdegree_data[1:6, ]

# What proportion have no outgoing ties?
targets$outdegree_data$mean_n[1] / nrow(nodes)
```

This distribution comes from a meta-analysis of empirical syringe-sharing studies —
it is the structural target our ERGM must reproduce.

---

## Social mixing: sex

```{r sex-mixing}
# Target: of all edges, what fraction involve male vs. female endpoints?
# And within those, how do they mix?
targets$sex_mixing_align_order

# Total edges target
targets$edges_target
```

---

## Social mixing: race/ethnicity

```{r race-mixing}
# 4x4 mixing matrix (White=1, Black=2, Hispanic=3, Other=4)
# Each cell = target number of directed edges
matrix(targets$target_race_num,
       nrow = 4, byrow = FALSE,
       dimnames = list(
         c("Wh","Bl","Hi","Ot"),  # sender
         c("Bl","Hi","Ot","Wh","Bl","Hi","Ot","Wh","Hi","Ot","Wh","Bl","Ot","Wh","Bl","Hi")
       ))
# Note: diagonal (Wh→Wh) excluded — it is the reference level (levels2 = -1)
```

What do these numbers tell us about racial homophily in this population?

---

## From summaries to targets

All of these — edges, sex mixing, age mixing, race mixing, degree distributions —
become the `target.stats` vector we pass to `ergm()` in Module 3.

```{r show-targets}
# What we'll use as fitting targets
cat("edges_target:", round(targets$edges_target), "\n")
cat("sex mixing targets:", round(targets$sex_mixing_align_order), "\n")
cat("age mixing targets:", round(targets$age_mixing_align_order), "\n")
```

---

## Exercise

1. Load the dataset and inspect the attribute distributions
2. What proportion of nodes are young (age < 26)? What proportion are non-Hispanic White?
3. Looking at the race mixing targets: which group shows the strongest within-group mixing?
4. What is the target mean out-degree? *(Hint: use `edges_target` and `nrow(nodes)`)*
5. Save your targets object — you'll use it in every subsequent module
