---
title: "Module 1: From Aggregate Summaries to Network Parameters"
subtitle: "SSC 2026 ERGM Workshop"
format: revealjs
---

## Learning Objectives

By the end of this module, you will be able to:

- Read in and summarize empirical network data
- Translate social mixing summaries into ERGM target parameters
- Specify the target network configuration for a synthetic population

---

## Setup

```{r setup}
library(network)
library(ergm)
library(dplyr)
library(readr)
library(ggplot2)

# Load the synthetic workshop dataset
nodes <- read_csv("../data/synthetic/nodes.csv")
edges <- read_csv("../data/synthetic/edges.csv")
```

---

## The Data: PWID Syringe-Sharing Network

- **n = 1,000 nodes** (people who inject drugs, Chicago metro area)
- **Attributes:** age, race/ethnicity, sex, geographic location
- **Ties:** syringe-sharing partnerships (undirected)

```{r inspect-data}
glimpse(nodes)
```

---

## Degree Distribution

How many syringe-sharing partners does each person have?

```{r degree-dist}
# TODO: compute degree distribution from edges
```

---

## Social Mixing: Race/Ethnicity

Do people tend to share syringes within or across racial/ethnic groups?

```{r mixing-matrix}
# TODO: compute mixing matrix
```

---

## Age Mixing

```{r age-mixing}
# TODO: compute age mixing summary
```

---

## Geographic Patterns

```{r geo-summary}
# TODO: summarize geographic distribution
```

---

## From Summaries to ERGM Targets

| Summary Statistic | ERGM Term |
|-------------------|-----------|
| Mean degree | `edges` |
| Degree distribution shape | `gwdegree` |
| Race/ethnicity homophily | `nodematch("race")` |
| Age mixing | `absdiff("age")` |
| Geographic distance | `dist` (custom term) |

---

## Exercise

1. Load the dataset
2. Compute the mixing matrix for race/ethnicity
3. Estimate the target number of edges from mean degree
4. Record your targets — you'll use them in Module 3
