---
title: "Module 3: Sequential ERGM Specification"
subtitle: "SSC 2026 ERGM Workshop"
format: revealjs
---

## Overview

**Goal:** Build a convergent ERGM through staged model fitting.

**Key ideas:**
- Each term has a substantive justification derived from Module 1 targets
- Add terms sequentially; simulate after each step to warm-start the next
- Fitting algorithm choice matters — and can itself be a diagnostic

---

## Fitting sequence

| Step | Terms added | Captures |
|------|-------------|----------|
| 1 | `edges + nodemix("sex")` | Total edges + sex mixing |
| 2 | `+ nodemix("young")` | Age mixing |
| 3 | `+ nodemix("race.num")` | Racial/ethnic mixing |
| 4 | `+ dnf(...)` | Geographic mixing *(Module 2)* |
| 5 | `+ odegree(0)` | Proportion of isolates |
| 6 | `+ odegree(0:1)` | Degree-0 and degree-1 nodes |
| 7 | `+ idegree(0:1)` | In-degree distribution |

---

## Fitting approach: default MCMLE

Start with the default algorithm — it works well for simpler models.

```{r fit-step1}
fit1 <- ergm(
  net ~
    edges +
    nodemix("sex", levels2 = -1),
  target.stats = c(edges_target, sex_mixing_align_order)
)
summary(fit1)
```

Does it converge? Check the MCMC diagnostics.

---

## When the default struggles

As the model grows more complex (more terms, harder geometry), MCMLE may fail to converge.

**Signs of trouble:**
- MCMC chains that don't mix
- Estimates that don't stabilize across iterations
- `ergm` warnings about non-convergence

---

## Remedy: Stochastic Approximation

SA is a more robust fitting algorithm — it was needed for the full HepCEP network.
Worth trying on the workshop data if the default struggles.

```{r fit-sa}
fit3 <- ergm(
  net ~
    edges +
    nodemix("sex",      levels2 = -1) +
    nodemix("young",    levels2 = -1) +
    nodemix("race.num", levels2 = -1),
  target.stats = c(
    edges_target,
    sex_mixing_align_order,
    age_mixing_align_order,
    target_race_num
  ),
  control = control.ergm(
    main.method = "Stochastic-Approximation",
    MCMLE.maxit = 500,
    MCMC.interval = 1e6,
    MCMC.samplesize = 1e6
  )
)
```

---

## Simulate after each step

```{r simulate}
net_next <- simulate(fit_current, nsim = 1)
```

**Why?** Warm-starting from a plausible network helps subsequent steps converge.

---

## Steps 5–7: Degree terms

```{r fit-degree}
# odegree(0): proportion with no outgoing ties
fit5 <- ergm(net5 ~ ... + odegree(0), target.stats = c(...))

# odegree(0:1): add degree-1 nodes
fit6 <- ergm(net6 ~ ... + odegree(0:1), target.stats = c(...))

# Add in-degree
fit7 <- ergm(net7 ~ ... + idegree(0:1) + odegree(0:1), ...)
```

---

## Discussion

- Did the default algorithm work, or did you need SA?
- How do simulated network properties change at each step?
- When should you stop adding terms?
