---
title: "Module 4: Assessing Failure Modes"
subtitle: "SSC 2026 ERGM Workshop"
format: revealjs
---

## Overview

**Goal:** Learn to distinguish technical convergence failures from substantive misspecification.

**Key ideas:**
- Not all problems look the same
- Some failures are algorithmic; others reflect a misspecified model
- Both matter, and they require different responses

---

## Two types of problems

| Type | Example | Fix |
|------|---------|-----|
| Technical | MCMC chains don't mix | Adjust control parameters, try SA |
| Degeneracy | Model implies near-empty or near-complete networks | Remove or reparameterize term |
| Misalignment | Simulated stats far from targets | Add missing term or revise target |

---

## Example: Degeneracy from `odegree(0:2)`

In the full pipeline, adding `odegree(0:2)` caused degeneracy.

```{r fit-degenerate}
# This model collapses to near-empty or near-complete networks
fit_degen <- ergm(
  net ~
    edges +
    nodemix("sex",      levels2 = -1) +
    nodemix("young",    levels2 = -1) +
    nodemix("race.num", levels2 = -1) +
    odegree(0:2),
  target.stats = c(...)
)
```

What does the simulated network look like?

---

## Diagnosing degeneracy

```{r diagnose}
net_degen <- simulate(fit_degen, nsim = 1)
summary(net_degen)

# Compare edge count to target
network.edgecount(net_degen)
edges_target
```

A degenerate network will have far too few or far too many edges.

---

## MCMC diagnostics

```{r mcmc-diag}
library(latticeExtra)
mcmc.diagnostics(fit_current)
```

**Look for:**
- Flat or drifting traces (non-mixing)
- Autocorrelation that doesn't decay
- Geweke statistic far from 0

---

## Non-convergence vs. poor alignment

**Non-convergence:** the algorithm failed to find the MLE
- Fix: more iterations, different algorithm, better starting network

**Poor alignment:** model converged, but simulated networks don't match targets
- Fix: review model specification, reconsider terms

---

## Exercise

1. Fit the degenerate model (`odegree(0:2)`) — what happens?
2. Run `mcmc.diagnostics()` on a converged vs. non-converged fit
3. Simulate from each and compare edge counts to target
