---
title: "Module 5: Simulation as Diagnostic"
subtitle: "SSC 2026 ERGM Workshop"
format: revealjs
---

## Overview

**Goal:** Use simulated network distributions to assess alignment with empirical targets.

**Key ideas:**
- A converged model is not necessarily a well-aligned model
- "Goodness of fit" means different things depending on your goal
- In an ABM pipeline, the targets are the ground truth — not a single observed network

---

## Two approaches to assessing fit

| Approach | What it compares to | When it fits |
|----------|---------------------|--------------|
| `gof()` | The fitted/observed network | Model fit to a single observed network |
| Manual simulation | Your empirical target statistics | Model fit to multi-source summary data |

**This pipeline:** targets come from empirical summaries (mixing matrices, degree distributions), not a single observed network. The manual approach is more directly interpretable.

---

## `gof()` — what it does

```{r gof}
# Standard ERGM goodness-of-fit
gof_result <- gof(fit_final)
plot(gof_result)
```

Compares simulated networks to the observed/fitted network on:
degree distribution, geodesic distances, edgewise shared partners.

Useful — but asks "does the model reproduce the observed network?"
not "does the model reproduce our empirical targets?"

---

## Manual simulation — the approach here

```{r simulate}
# Simulate many networks from the fitted model
nets <- simulate(fit_final, nsim = 100)
```

Then compute the same statistics you derived in Module 1 for each simulated network and compare to targets.

---

## Example: Out-degree distribution

```{r violin-degree}
library(ggplot2)

sim_odeg0 <- sapply(nets, function(net) {
  sum(degree(net, cmode = "outdegree") == 0) / network.size(net)
})

ggplot(data.frame(val = sim_odeg0), aes(x = "", y = val)) +
  geom_violin() +
  geom_hline(yintercept = target_odeg0, color = "red", linewidth = 1) +
  labs(y = "Proportion with out-degree 0", x = "") +
  theme_minimal()
```

Red line = empirical target. Does the simulated distribution cover it?

---

## Example: Racial mixing

```{r violin-mixing}
# For each simulated network, extract mixing matrix
# Compare each cell to the corresponding target
# TODO: example code
```

---

## What good alignment looks like

- Target falls within the bulk of the simulated distribution
- No systematic bias (target not always at one tail)
- Tight distributions = stable, low-variance model

---

## What poor alignment tells you

- Target consistently outside the simulated distribution → misspecification
- Which statistics are off? That points to which terms need revision
- Loop back to Module 3 with a revised model

---

## Exercise

1. Simulate 100 networks from your fitted model
2. Plot violin plots for: out-degree distribution, racial mixing, edge count
3. Identify any statistics where alignment is poor
4. What would you change in the model to address it?
