---
title: "Module 6: What Do You Do With a Simulated Network?"
subtitle: "SSC 2026 ERGM Workshop"
format: revealjs
---

## Overview

**Goal:** Situate the ERGM pipeline in broader use contexts.

**Key ideas:**
- The simulated network is a product in its own right — and an input to further modeling
- Multiple simulations matter whether the network is the end goal or a stepping stone
- Different frameworks integrate network modeling and simulation differently

---

## Networks as an end product

Simulated networks can be the primary deliverable:
- Study structural properties of a population (connectivity, clustering, mixing)
- Compare network configurations under different assumptions
- Assess how well a generative model reproduces observed structure

In this case, the ensemble of simulated networks *is* the result — and the diagnostic work in Module 5 is the analysis.

---

## Networks as ABM input (this use case)

In HepCEP, simulated networks feed an agent-based model:
- Nodes become agents with attributes
- Edges define partnership structure at model initialization
- Network uncertainty propagates into ABM output uncertainty

Running the ABM across an ensemble of networks quantifies how much
network variation matters for intervention outcomes.

---

## Extract vertex attributes

```{r vertex-attrs}
attrs <- data.frame(
  id       = network.vertex.names(net),
  sex      = get.vertex.attribute(net, "sex"),
  young    = get.vertex.attribute(net, "young"),
  race_num = get.vertex.attribute(net, "race.num"),
  chicago  = get.vertex.attribute(net, "chicago")
)
write.csv(attrs, "out/vertex_attributes.csv", row.names = FALSE)
```

---

## Extract edgelist

```{r edgelist}
el <- as.edgelist(net)
colnames(el) <- c("from", "to")
write.csv(el, "out/edgelist.csv", row.names = FALSE)
```

---

## Propagating network uncertainty into the ABM

```{r ensemble}
for (i in seq_along(nets)) {
  el_i <- as.edgelist(nets[[i]])
  write.csv(el_i, paste0("out/edgelist_", i, ".csv"), row.names = FALSE)
}
```

Run the ABM once per network — variation in outcomes reflects network uncertainty.

---

## Integrated frameworks

Some frameworks handle network modeling and epidemic simulation together:

- **EpiModel** — fits ERGMs and runs epidemic simulations in one pipeline; dynamic network models evolve over time
- **HepCEP** — separates network generation (this pipeline) from ABM; allows richer agent behavior and intervention logic

**Trade-off:** integration is convenient; separation gives more control.

---

## The full pipeline

```
Empirical data
  → Module 1: Target statistics
  → Module 2: Geographic term (Jonathan)
  → Module 3: Fitted ERGM
  → Module 4: Convergence checks
  → Module 5: Simulation + alignment
  → Module 6: Use — structural analysis or ABM input
```

---

## Discussion

- Is the network an end product or a means to an end in your work?
- How does your choice of framework shape what questions you can ask?
- What assumptions embedded in the network carry forward into downstream analysis?
