---
title: "From Aggregate Summaries to Network Targets"
subtitle: "SSC 2026 ERGM Workshop · Module 2 — exercise"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  eval: false     # run the chunks yourself in RStudio / Posit Cloud
  echo: true
  warning: false
  message: false
---

> **Live concept.** Turn a degree distribution into an edge count. Full worked
> version: `R/01-targets.R`.

## Setup

Loads the synthetic data and the `targets` list (prints a short setup line).

```{r}
#| label: setup
source(here::here("R", "00-setup.R"))
```

## Run this: in-edges from the in-degree distribution

`indegree_data` has one row per degree: `in_degree`, and `num_n` = the number of
people reporting that in-degree.

```{r}
#| label: in-edges
indeg <- targets$indegree_data
indeg
```

Every tie is one partner-slot for someone, so the total number of in-edges is the
sum, over degrees, of `degree * (# people at that degree)`:

```{r}
#| label: in-edges-sum
inedges_target <- sum(indeg$in_degree * indeg$num_n)
inedges_target                      # ~ 630.8
```

## Your turn: repeat for out-degree, then reconcile

Mirror the in-edges calculation for the **out-degree** distribution, then average
the two into a single edge target. In- and out-edges count the *same* ties, so they
should agree; they don't exactly (each is estimated separately), so we average them.

```{r}
#| label: your-turn
outdeg <- targets$outdegree_data

# TODO: mirror the in-edges line above, for the OUT-degree distribution.
# outedges_target <- sum(outdeg$________ * outdeg$________)

# TODO: average the two into one edge target (~ 711).
# edges_target <- mean(c(inedges_target, outedges_target))
# edges_target
```

::: {.callout-note}
## Try at home

a. Turn mixing **proportions** into counts. A sex-mixing cell is, e.g.,
   `edges_target * male_share * male_within`. Then match the cell **order** to
   `summary(net ~ nodemix("sex", levels2 = -1))`.
b. See how `R/01-targets.R` assembles the full target vector
   `c(edges_target, sex_target, age_target, race_target)` and length-checks it
   against the number of ERGM terms.
:::

::: {.callout-tip collapse="true"}
## Solution

```r
outdeg <- targets$outdegree_data
outedges_target <- sum(outdeg$out_degree * outdeg$num_n)

edges_target <- mean(c(inedges_target, outedges_target))   # ~ 711
edges_target
```
:::
