---
title: "Sequential ERGM Specification"
subtitle: "SSC 2026 ERGM Workshop · Module 3 — exercise"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  eval: false     # run the chunks yourself in RStudio / Posit Cloud
  echo: true
  warning: false
  message: false
---

> **Live concept.** Fit a block, **simulate** it, and use that network to warm-start
> the next, harder fit. Full staged sequence: `R/02-sequential.R`.

## Setup

Loads `net`, `ts_mix`, and two helpers, `odeg_target(k)` / `ideg_target(k)`. These
return the **target counts** for the degree terms — the number of nodes with out-
(or in-) degree `k`, read straight from the degree distribution you built in Module 2.
So `odeg_target(0)` is the count of people with out-degree 0, and `odeg_target(0:1)`
the counts for degrees 0 and 1. You pass these alongside `ts_mix` as `target.stats`.

```{r}
#| label: setup
source(here::here("R", "00-setup.R"))
```

The dyad-independent mixing block (edges + attribute mixing + geography):

```{r}
#| label: mixing-formula
f_mix <- net ~ edges +
  nodemix("sex",      levels2 = -1) +
  nodemix("young",    levels2 = -1) +
  nodemix("race.num", levels2 = -1) +
  nodematch("chicago")
```

## Controls

The mixing block is dyad-independent and fits quickly. The degree terms are
dyad-dependent. Under MCMLE defaults the sampler keeps enlarging and can run for a
long time at n = 1000. Therefore, we instead fit these with **Stochastic-Approximation** (SA), which is
based on **moment-matching** (Robbins–Monro).^[Snijders (2002), *Journal of Social Structure* 3(2), 1–40.]  

SA works by simulating networks from the current fit, comparing the simulated
statistics to the targets, and **nudging all the coefficients jointly**. This process is repeated until the simulated statistic vector is close to the targets.

In our study, SA often seemed to converge more reliably for dyad-dependent networks, especially at the larger scale.

We set `MCMC.interval`/`samplesize = 4096`. Each SA step **simulates** networks
from the current fit by MCMC to estimate the expected statistics, so these `MCMC.*`
values govern *that* sampler (under SA they set `SA.interval`/`SA.samplesize`). A larger
sample gives tighter, less-autocorrelated draws, improving the quality of the simulated network (see Modules 4–5 on assessment of quality).

```{r}
#| label: controls
# ctrl: light, capped MCMLE -- we DON'T use it in the live fits; it's here for the
# take-home, where you swap it in to watch a degree fit struggle under MCMLE.
ctrl <- control.ergm(MCMC.interval = 512, MCMC.samplesize = 512,
                     MCMC.effectiveSize = NULL, MCMLE.maxit = 30)

# sa_control: what the degree fits below actually use.
sa_control <- control.ergm(main.method = "Stochastic-Approximation",
                           MCMC.interval = 4096, MCMC.samplesize = 4096)
```

## Run this: fit the mixing block, then warm-start

```{r}
#| label: fit-mix
# 1. Fit the mixing block to its targets.
fit_mix <- ergm(f_mix, target.stats = ts_mix, control = sa_control, eval.loglik = FALSE)

# 2. Simulate one network from that fit -> the "warm start" for the next fit.
net_warm <- simulate(fit_mix, nsim = 1)

# 3. Add an out-degree term, fitting FROM the warm start (net_warm on the LHS).
fit5 <- ergm(update(f_mix, net_warm ~ . + odegree(0)),
             target.stats = c(ts_mix, odeg_target(0)),
             control = sa_control, eval.loglik = FALSE)
summary(fit5)
```

## Your turn: add `odegree(0:1)` the same way

Warm-start again from `fit5`, then add `odegree(0:1)` (targets for degrees 0 and 1).

```{r}
#| label: your-turn
# TODO: net_warm2 <- simulate(fit5, nsim = 1)
# TODO: fit6 <- ergm(update(f_mix, net_warm2 ~ . + odegree(0:1)),
#                    target.stats = c(ts_mix, odeg_target(0:1)),
#                    control = sa_control, eval.loglik = FALSE)
# summary(fit6)
```

::: {.callout-note}
## Try at home

a. Refit `fit5` with `control = ctrl` (light, capped MCMLE) instead of `sa_control`
   and watch the difference — it runs slower and may not match the degree targets
   within the capped iterations. That contrast is exactly what SA solves here.
b. The hard step: add BOTH `idegree(0:1) + odegree(0:1)`. Under ergm's default
   (MCMLE) it fits slowly; switching to SA is far more efficient:
   `control = control.ergm(main.method = "Stochastic-Approximation")`. Full staged
   sequence + the SA switch: `R/02-sequential.R`.
:::

::: {.callout-tip collapse="true"}
## Solution

```r
net_warm2 <- simulate(fit5, nsim = 1)
fit6 <- ergm(update(f_mix, net_warm2 ~ . + odegree(0:1)),
             target.stats = c(ts_mix, odeg_target(0:1)),
             control = sa_control, eval.loglik = FALSE)
summary(fit6)
```
:::