---
title: "Assessing Failure Modes"
subtitle: "SSC 2026 ERGM Workshop · Module 4 — exercise"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  eval: false     # run the chunks yourself in RStudio / Posit Cloud
  echo: true
  warning: false
  message: false
---

> **Live concept.** Push a degree term too far and watch the fit blow up. Work top
> to bottom; fill in the `# TODO` chunk. Full worked version: `R/04-failure-modes.R`.

## Setup

```{r}
#| label: setup
source(here::here("R", "00-setup.R"))
```

The dyad-independent mixing block (same as Module 3):

```{r}
#| label: mixing-formula
f_mix <- net ~ edges +
  nodemix("sex",      levels2 = -1) +
  nodemix("young",    levels2 = -1) +
  nodemix("race.num", levels2 = -1) +
  nodematch("chicago")
```

## Controls

Two controls: `mcmle_ctrl` keeps MCMLE for the degenerate fit so it surfaces the
degeneracy **error** (the lesson), with iterations capped so it gives up quickly.
`sa_control` is for the non-degenerate "your turn" fit — degree terms fit reliably
under Stochastic-Approximation (see Module 3 / `R/02-sequential.R`).

```{r}
#| label: controls
mcmle_ctrl <- control.ergm(MCMLE.maxit = 20)
sa_control <- control.ergm(main.method = "Stochastic-Approximation",
                           MCMC.interval = 4096, MCMC.samplesize = 4096)
```

## Run this: trigger degeneracy

`odegree(0:2)` fits fine at n = 1000; `odegree(0:3)` is **degenerate** — ergm errors
out. Read the message: it flags simulated networks nothing like the data
(*"...exceeds that in the observed by a factor of more than 20..."*). Watch also for
a warning that the model is **nonidentifiable** — `edges` and the full `odegree(0:3)`
set are linearly dependent
(`CONSTANT = edges + 4*odegree0 + 3*odegree1 + 2*odegree2 + odegree3`), a second
reason this specification breaks.

```{r}
#| label: trigger-degeneracy
ergm(update(f_mix, ~ . + odegree(0:3)),
     target.stats = c(ts_mix, odeg_target(0:3)),
     control = mcmle_ctrl, eval.loglik = FALSE)
```

## Your turn: back off to `odegree(0:2)`

Re-fit with `odegree(0:2)` under SA; it converges and simulates a sensible edge count.

```{r}
#| label: your-turn
# TODO: fit_ok <- ergm(update(f_mix, ~ . + odegree(0:2)),
#                      target.stats = c(ts_mix, odeg_target(0:2)),
#                      control = sa_control, eval.loglik = FALSE)
# TODO: network.edgecount(simulate(fit_ok, nsim = 1))    # near the target now?
```

::: {.callout-note}
## Try at home

a. Read the sampler: `mcmc.diagnostics(fit_ok)`. Healthy traces look like flat
   "fuzzy caterpillars"; drifting traces signal non-convergence.
b. `gof(fit_ok, GOF = ~ idegree + odegree - model)` — do the degree distributions
   line up? (Full version: `R/04-failure-modes.R`.)
:::

::: {.callout-tip collapse="true"}
## Solution

```r
fit_ok <- ergm(update(f_mix, ~ . + odegree(0:2)),
               target.stats = c(ts_mix, odeg_target(0:2)),
               control = sa_control, eval.loglik = FALSE)
network.edgecount(simulate(fit_ok, nsim = 1))    # near the target now
```
:::
