---
title: "Simulation as Diagnostic"
subtitle: "SSC 2026 ERGM Workshop · Module 5 — exercise"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  eval: false     # run the chunks yourself in RStudio / Posit Cloud
  echo: true
  warning: false
  message: false
---

> **Live concept.** Simulate an ensemble and check whether a target lands inside the
> simulated spread. Work top to bottom; fill in the `# TODO` chunk. Full worked
> versions: `R/05a-simulation.R`, `R/05b-export-abm.R`.

## Setup

```{r}
#| label: setup
source(here::here("R", "00-setup.R"))
```

The mixing block (same as Module 3):

```{r}
#| label: mixing-formula
f_mix <- net ~ edges +
  nodemix("sex",      levels2 = -1) +
  nodemix("young",    levels2 = -1) +
  nodemix("race.num", levels2 = -1) +
  nodematch("chicago")
```

## Controls and a fitted model

Degree terms fit far faster under Stochastic-Approximation at n = 1000 (MCMLE
defaults run long). Sampling level matters here too: too-low MCMC leaves consecutive
draws autocorrelated, so the simulated intervals below come out misleadingly narrow;
4096 gives honest intervals in seconds (see `R/02-sequential.R`).

```{r}
#| label: fit
sa_control <- control.ergm(main.method = "Stochastic-Approximation",
                           MCMC.interval = 4096, MCMC.samplesize = 4096)

# A fitted model to simulate from (mixing block + out-degree; fits in seconds).
fit <- ergm(update(f_mix, ~ . + odegree(0:1)),
            target.stats = c(ts_mix, odeg_target(0:1)),
            control = sa_control, eval.loglik = FALSE)
```

## Run this: simulate an ensemble, check ONE statistic

```{r}
#| label: ensemble
sims <- simulate(fit, nsim = 100)

# Count out-degree-0 nodes in each of the 100 simulated networks:
sim_odeg0    <- sapply(seq_along(sims), function(i) summary(sims[[i]] ~ odegree(0)))
target_odeg0 <- odeg_target(0)

# Is the target inside the middle 95% of the simulated distribution?
quantile(sim_odeg0, c(.025, .975))
target_odeg0
target_odeg0 >= quantile(sim_odeg0, .025) & target_odeg0 <= quantile(sim_odeg0, .975)
```

## Your turn: check a different statistic

Repeat for **out-degree-1**. Is *its* target, `odeg_target(1)`, inside the spread?

```{r}
#| label: your-turn
# TODO: sim_odeg1 <- sapply(seq_along(sims), function(i) summary(sims[[i]] ~ odegree(1)))
# TODO: quantile(sim_odeg1, c(.025, .975)); odeg_target(1)
```

::: {.callout-note}
## Try at home

a. Violin plot of every statistic vs. its target: `R/05a-simulation.R`.
b. Export a simulated network for an ABM (edgelist + vertex table):
   `R/05b-export-abm.R`.
:::

::: {.callout-tip collapse="true"}
## Solution

```r
sim_odeg1    <- sapply(seq_along(sims), function(i) summary(sims[[i]] ~ odegree(1)))
target_odeg1 <- odeg_target(1)

quantile(sim_odeg1, c(.025, .975))
target_odeg1
target_odeg1 >= quantile(sim_odeg1, .025) & target_odeg1 <= quantile(sim_odeg1, .975)
```
:::
