---
title: "Geographic Extensions"
subtitle: "SSC 2026 ERGM Workshop · Module 6 — exercise"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  eval: false     # run the chunks yourself in RStudio / Posit Cloud
  echo: true
  warning: false
  message: false
---

> **Live concept.** `nodematch` only asks whether two labels match; `dnf` also asks
> how far the tie went. Check both on a network small enough to verify by hand. No
> fitting here, so the whole exercise runs in seconds. Full walkthrough:
> `modules/06-geographic-extensions.qmd` (Module 6 has no `R/` script).

## Setup

`ergm.userterms.hepcep` is pre-installed in the Posit Cloud workspace.

```{r}
#| label: setup
library(ergm)                     # attaches `network` too
library(ergm.userterms.hepcep)    # custom geographic terms
```

Three Hyde Park landmarks, three directed ties. Nodes 1 and 2 are category 1, node 3
is category 2.

```{r}
#| label: build-network
m <- matrix(c(1,2, 2,3, 3,1), byrow = TRUE, ncol = 2)   # 1->2, 2->3, 3->1
g <- as.network(m, matrix.type = "edgelist", directed = TRUE)

g %v% "lat"  <- c(41.796027, 41.790636, 41.768876)   # Promontory Pt, MSI, South Shore
g %v% "lon"  <- c(-87.577310, -87.582482, -87.562375)
g %v% "cat1" <- c(1, 1, 2)

# Tie distances, worked out from those coordinates:
#   1->2 = 0.73 km     2->3 = 2.92 km     3->1 = 3.24 km
```

## Run this: what `nodematch` cannot see

```{r}
#| label: nodematch
summary(g ~ nodematch("cat1"))    # 1 -- the one within-category tie, 1->2
```

That `1` would be unchanged if `1->2` were 300 km long. `dnf` cross-classifies each
tie by the **source** node's category and by whether the tie falls under that
category's threshold. `base` says which of the `2n` cells to omit as a reference.

```{r}
#| label: dnf-both-bases
summary(g ~ dnf(by = "cat1", thresholds = c(2, 2), base = 4))
#> dnf.cat1.1.n dnf.cat1.1.f dnf.cat1.2.n
#>            1            1            0

summary(g ~ dnf(by = "cat1", thresholds = c(2, 2), base = 1))
#> dnf.cat1.1.f dnf.cat1.2.n dnf.cat1.2.f
#>            1            0            1
```

Reading both calls: `1.n = 1`, `1.f = 1`, `2.n = 0`, `2.f = 1`. Only `1->2` beat the
2 km threshold. The four cells sum to 3, the edge count — every tie lands in exactly
one cell. That sum check is the cheapest correctness test for a term like this.

```{r}
#| label: edge-count
network.edgecount(g)
```

## Your turn: move one threshold

Tighten category 1 to 0.1 km and leave category 2 at 2 km. **Predict** which cells
change before you run it: with a 0.1 km cutoff, is `1->2` still "near"?

```{r}
#| label: your-turn
# TODO: summary(g ~ dnf(by = "cat1", thresholds = c(0.1, 2), base = 4))
# TODO: summary(g ~ dnf(by = "cat1", thresholds = c(0.1, 2), base = 1))
# Do the four cells still sum to 3?
```

::: {.callout-note}
## Try at home

a. Swap the stand-in for the real term on the n = 1000 network and count statistics.
   Run `source(here::here("R", "00-setup.R"))` first, then compare the mixing block
   ending in `nodematch("chicago")` (23 stats) against one ending in
   `dnf(by = "chicago", thresholds = c(2, 2))` (25 stats).
b. Refit that mixing block with the four near/far targets (`.457, .229, .163, .151`
   of `edges_target`, from the survey summaries). Does it still converge?
c. The sibling term: `summary(g ~ dist(1:3))` bands ties by distance and ignores
   category entirely. Which question does each term answer that the other can't?
:::

::: {.callout-tip collapse="true"}
## Solution

With a 0.1 km cutoff, `1->2` (0.73 km) is no longer "near" — it moves from `1.n` to
`1.f`. So `1.n` goes `1 -> 0` and `1.f` goes `1 -> 2`; the four cells still sum to 3.

```r
summary(g ~ dnf(by = "cat1", thresholds = c(0.1, 2), base = 4))
#> dnf.cat1.1.n dnf.cat1.1.f dnf.cat1.2.n
#>            0            2            0

summary(g ~ dnf(by = "cat1", thresholds = c(0.1, 2), base = 1))
#> dnf.cat1.1.f dnf.cat1.2.n dnf.cat1.2.f
#>            2            0            1
```

Full set: `1.n = 0`, `1.f = 2`, `2.n = 0`, `2.f = 1` — sum = 3 = `network.edgecount(g)`.
:::
