---
title: "Module 1: Introduction to ERGMs"
subtitle: "SSC 2026 ERGM Workshop"
format: revealjs
---

## The relational perspective

- People influence each other
- People who share attributes are more likely to be connected
- When studying health, what is happening *around people* needs explicit attention

**Today's context:** syringe-sharing partnerships among people who inject drugs (PWID)
in the Chicago metro area — a network that drives HCV and HIV transmission

---

## Nodes, edges, and network structure

- **Nodes** represent individual entities (here: people)
- **Edges** represent connections (here: syringe-sharing partnerships)
- **Structure** = the pattern of who is connected to whom

```{r setup}
library(network)
library(ergm)
library(dplyr)
library(readr)
library(here)

nodes   <- read_csv(here("data", "synthetic", "nodes.csv"))
targets <- readRDS(here("data", "synthetic", "targets.rds"))

# Initialize a directed network with node attributes
net <- network.initialize(n = nrow(nodes), directed = TRUE)
set.vertex.attribute(net, colnames(nodes), nodes)
net
```

---

## Our research question

**What social processes generate the observed syringe-sharing structure
in this population?**

- Do people share syringes within racial/ethnic groups? With similar-age partners?
- How many sharing partners does a typical person have?
- Does geography shape who shares with whom?

We need a model that can represent these processes formally and estimate their effects.

---

## What is an ERGM?

An **Exponential Random Graph Model** asks:

> Given a set of nodes with known attributes, what is the probability of
> observing a particular network?

ERGMs are operationalized like logistic regression:

- **Outcome:** does a tie exist between nodes *i* and *j*? (0/1)
- **Predictors:** network statistics capturing structural and attribute-based processes
- **Coefficients:** log-odds of a tie existing

$$\text{logit}(P(y_{ij} = 1)) = \theta_1 g_1(\mathbf{y}, \mathbf{X}) + \theta_2 g_2(\mathbf{y}, \mathbf{X}) + \ldots$$

---

## The simplest ERGM: edges only

The `edges` term is the intercept — it captures the baseline density of the network.

```{r fit-edges}
fit_edges <- ergm(net ~ edges,
                  target.stats = targets$edges_target)
summary(fit_edges)
```

- The edges coefficient = log-odds of a random tie existing
- Convert to probability: `plogis(coef(fit_edges)["edges"])`
- What does this probability tell us?

---

## Adding social processes: sex mixing

Do men and women share syringes assortatively?

```{r fit-sex}
fit_sex <- ergm(net ~ edges +
                  nodemix("sex", levels2 = -1),
                target.stats = c(targets$edges_target,
                                 targets$sex_mixing_align_order))
summary(fit_sex)
```

- Positive coefficient → ties more likely between that combination
- How does this compare to the null (random) mixing expectation?

---

## Interpretation

The `nodemix` coefficients are log-odds *relative to the reference category*.

```{r interpret}
# Convert to odds ratios
exp(coef(fit_sex))
```

- Which sex combination is the reference?
- How much more (or less) likely are within-sex ties vs. the reference?

---

## Adding more processes: age and race

```{r fit-full-mixing}
fit_mixing <- ergm(net ~ edges +
                     nodemix("sex",      levels2 = -1) +
                     nodemix("young",    levels2 = -1) +
                     nodemix("race.num", levels2 = -1),
                   target.stats = c(targets$edges_target,
                                    targets$sex_mixing_align_order,
                                    targets$age_mixing_align_order,
                                    targets$target_race_num))
summary(fit_mixing)
```

Each term answers a specific substantive question about tie formation.

---

## Simulating from a fitted model

Once a model is fit, we can **generate networks** from it.

```{r simulate}
net_sim <- simulate(fit_mixing, nsim = 1)
net_sim

# Compare edge count to target
network.edgecount(net_sim)
targets$edges_target
```

This is the key idea behind the rest of the workshop:
*use simulation to assess whether the model reproduces what we care about.*

---

## What makes a good model?

A converged ERGM that simulates networks matching the target structure.

| Check | Tool |
|-------|------|
| Did the algorithm converge? | MCMC diagnostics |
| Do simulated networks match targets? | Violin plots vs. target statistics |
| Is the model missing a process? | Substantive judgment |

Modules 3–5 work through each of these in depth.

---

## The workshop pipeline

```
Empirical data
  → Module 1: Understand ERGMs, fit simple models    ← you are here
  → Module 2: Geographic mixing term (Jonathan)
  → Module 3: Sequential ERGM fitting
  → Module 4: Diagnosing failure modes
  → Module 5: Simulation as diagnostic
  → Module 6: What do you do with a simulated network?
```

---

## Exercise

1. Load the dataset and initialize the network object
2. Fit the `edges`-only model — what is the implied tie probability?
3. Add `nodemix("sex", levels2 = -1)` — which sex combination shows the strongest tendency to share?
4. Simulate one network from your fitted model — how close is the edge count to the target?
